Asymmetric relationships and their impact on coexistence: Supporting information

### 2023-01-27 paper/Paper.Rnw
working on revisions to Methods and associated sections of sup mat

%Lottery model and results based on it
To further demonstrate the plausibility of ATAs altering coexistence between species, we carried out 
simple simulations using a two-species model, the well-known lottery model. The simulations described in this
paragraph are sufficient to demonstrate that ATAs influence coexistence in some manner; the
precise nature of that influence is the subject of the rest of the paper.
The lottery model is as follows.
Letting $N_i(t)$ denote the population density of species $i=1,2$ at time $t$, and defining $N = N_1(t)+N_2(t)$,
model equations are
\begin{equation}
N_i(t+1)=(1-\delta)N_i(t)+\delta N \frac{B_i(t)N_i(t)}{B_1(t)N_1(t)+B_2(t)N_2(t)}  \label{model_eq}
\end{equation}
for $i=1,2$. Here, $\delta$ is a mortality rate, and $B_i(t)$ is the fecundity of species $i$ at time $t$. 
The model postulates that individuals die at rate $\delta$ at each time step, and are replaced by juveniles 
in proportion to the reproductive outputs of the two species that year. Note that $N$ is constant through time. 
For each $i$, we assume for simplicity that the random variables $B_i(t)$ are independent and identically 
distributed (iid) through time. In three distinct similations (Fig. \ref{fig:pedagog1}d-f), 
$B_1$ and $B_2$ were, respectively, 
left- (Fig. \ref{fig:pedagog1}a), right- (Fig. \ref{fig:pedagog1}c) and symetrically (Fig. \ref{fig:pedagog1}b)
tail associated, while the same marginal distributions for these fecundities were used in all simulations (see Methods
for details). Thus, the model setup was identical in the three simulations except for different ATAs. 
ATAs substantially influenced aspects of species coexistence (Fig. \ref{fig:pedagog1}d-f).

@@

%Goal para
We previously demonstrated that ATAs are common in nature can influence ecological phenomena
\citep{Ghosh_2020_AER, Ghosh_2020_Eco, Ghosh_2020_ecolevol, Ghosh_2021}; and our arguments above
indicate that ATAs may influence species coexistence. Thus, we here seek to quantify the consequences 
of ATAs for competition, coexistence, and the maintenance of biodiversity, by means of the 
following specific goals. 
G1) To augment MCT to show formally how ATAs play a role in coexistence. 
MCT decomposes a measure of coexistence into components due to each of several mechanisms. 
We address how ATAs contribute to these components. 
We examine mechanisms of coexistence that apply in models exhibiting solely temporal variation; 
spatial variation can be considered in future work.
G2) To explore, using the simple lottery model, the relative magnitudes of the contributions to coexistence that come
from ATAs compared to other, previously known mechanisms of coexistence; and to determine whether species may 
sometimes coexist with ATAs but not without, or vice versa. This is to help 
illuminate whether ATA contributions to coexistence and biodiversity maintenance
are likely to be negligible or meaningfully large. 
G3) To provide an empirical example of species coexistence which highlights the role ATAs can play.
In the Discussion, we consider whether climate change may alter
ATAs of environmental variables and thereby influence competition and coexistence.
Overall, our study presents a new mechanism of species coexistence
and a means of understanding its theoretical and empirical importance.

@@

We here pursue goal G1 from the Introduction: to augment MCT to show formally how ATAs play a role in coexistence. 
We develop our new theory for a general two-species model, later applying it to specific scenarios.
MCT and its recent computational extensions \citep{Ellner_2016,Ellner_2019} 
quantify the contributions of 
multiple mechanisms to coexistence. But we show below
that, for purely temporal variation, only one mechanism relates to tail associations: storage effects. 
So we begin by defining storage effects. Our mathematical notation parallels
that of \cite{Ellner_2016,Ellner_2019}. It is assumed that the growth rate $r_i(t)=\ln(N_i(t+1)/N_i(t))$ 
of species $i$ can be written as an increasing function of
an environment-dependent factor $E_i(t)$ and as a decreasing function of a quantity representing
competitive pressure on species $i$, $C_i(t)$. We also assume, for simplicity, 
that the $E_i(t)$ are iid through time, and likewise for the
$C_i(t)$. Though the iid assumption can be circumvented, we make the assumption because it simplifies the
presentation of new ideas.

@@

The storage effect for species $i$ is the contribution of covariation between $E$ and 
$C$ to a difference between rare and common species mean growth rates, when $i$ is rare.
Specifically, storage effects for species $i$ are 
\begin{equation}
\Delta_i^{(EC)} = (\olr_{i\bs i} - q_{ij} \olr_{j \bs i})-(\olrs_{i \bs i}-q_{ij} \olrs_{j \bs i}),\label{eq:storage_effects_gen}
\end{equation}
where: $\olr_{i \bs i}$ refers to the mean ``growth when rare'' [GWR] of species $i$, that is, the mean growth rate of 
species $i$ when it is at negligible abundance and species $j$ (the other species) is at stochastic steady state;
$\olr_{j \bs i}$ refers to the mean steady-state 
growth rate of species $j$ when species $i$ is absent or at negligible abundance (this is $0$, \emph{a priori});
and $\olrs_{i \bs i}$ and $\olrs_{j \bs i}$ are analogous quantities, but for an alternative modelling scenario, identical except that
influences of environment and competition on either species have been rendered independent.  
A subscript 
$i \bs i$ always indicates a quantity computed for species $i$ when that species is at negligible abundance, and 
a subscript $j \bs i$ indicates a quantity computed for $j$ when $i$ is absent.
The \emph{scaling factor}
$q_{ij}$ measures the relative sensitivities to competition of the two species, see \cite{Chesson_1994},
\cite{Ellner_2016} and SI section \ref{SI-SIsect:scaling_factors} for the two
alternative definitions of $q_{ij}$ that we use. 
Since it was assumed that $r_i=r_i(E_i,C_i)$, we can write $\olr_{i \bs i}=\E[r_i(E_i,C_{i \bs i})]$, where
$\E(\cdot)$ denotes expected value just as does an overbar, but we use $\E$ for longer expressions.
Also, $\olr_{j \bs i}=\E[r_j(E_j,C_{j \bs i})]$.
An expression for $\olrs_{i \bs i}$ is obtained from $\olr_{i \bs i}$ by replacing the bivariate random 
variable $(E_i,C_{i \bs i})$ by a random variable with the same marginal distributions, but independent components, 
$(E_i^\shp,C_{i \bs i}^\shp)$, so $\olrs_{i \bs i}=\E[r_i(E_i^\shp,C_{i \bs i}^\shp)]$.
Likewise, $\olrs_{j \bs i}=\E[r_j(E_j^\shp,C_{j \bs i}^\shp)]$.

@@

The other mechanisms in MCT that apply in our modelling context of purely temporal variation
make no reference to covariation between quantities, whereas ATAs are 
defined in a covariation context. Therefore, storage effects are the only mechanism of coexistence
for which ATAs are relevant. Mathematical details supporting this claim are in SI section 
\ref{SI-SIsect:other_mechanisms}, but we here provide a conceptual understanding. \cite{Ellner_2016} and
\cite{Ellner_2019} provide a decomposition of the coexistence metric $\olr_{i \bs i} -q_{ij}\olr_{j \bs i}$,
which equals GWR of species $i$. The decomposition, which we use below, is similar to and motivated by the 
original decomposition of \cite{Chesson_1994}; the decomposition is how the mechanisms of 
coexistence operating in a given case are illuminated. The decomposition is 
\begin{equation}
\olr_{i \bs i} -q_{ij}\olr_{j \bs i}=\Delta_i^0+\Delta_i^E+\Delta_i^C+\Delta_i^{(E\shp C)}+\Delta_i^{(EC)},\label{eq:previous_decomp}
\end{equation}
where: $\Delta_i^0$ is a baseline;
$\Delta_i^E$ is a contribution to coexistence of environmental variation;
$\Delta_i^C$ is a contribution to coexistence of variation in competition;
$\Delta_i^{(E\shp C)}$ is a contribution to coexistence of having simultaneous variation in both environment and competition, but not including the effects 
of covariation in these quantities; and
$\Delta_i^{(EC)}$ is a contribution to coexistence of $EC$ covariation itself -- storage effects. 
Only storage effects involve $EC$ covariation,
so the only mechanism of coexistence influenced by ATA between $E$ and $C$ is storage effects. Combining our decomposition 
(\ref{eq:storage_effects_further}) with the decomposition (\ref{eq:previous_decomp}) of \cite{Ellner_2019} gives 
\begin{equation}
\olr_{i \bs i} -q_{ij}\olr_{j \bs i}=\Delta_i^0+\Delta_i^E+\Delta_i^C+\Delta_i^{(E\shp C)}+\Delta_i^{[E \nat C]} + \Delta_i^{[EC]},\label{eq:full_decomp}
\end{equation}
which is the extension of MCT which was goal G1 of the Introduction. One typically (but see below) must examine this quantity for each species, $i$,
to obtain complete information about coexistence. To pursue goal G2, below, we will quantify
the magnitude of the new ATA term, $\Delta_i^{[EC]}$, relative to the magnitudes of the other terms for the 
lottery model described in the Introduction.

@@

>>>>

%Lottery model and results based on it
To further demonstrate the plausibility of ATAs altering coexistence between species, we carried out 
simple simulations using a two-species model, the well-known lottery model. The simulations described in this
paragraph are sufficient to demonstrate that ATAs influence coexistence in some manner; the
precise nature of that influence is the subject of the rest of the paper.
The lottery model is as follows.
Letting $N_i(t)$ denote the population density of species $i=1,2$ at time $t$, and defining $N = N_1(t)+N_2(t)$,
model equations are
\begin{equation}
N_i(t+1)=(1-\delta)N_i(t)+\delta N \frac{B_i(t)N_i(t)}{B_1(t)N_1(t)+B_2(t)N_2(t)}  \label{model_eq}
\end{equation}
for $i=1,2$. Here, $\delta$ is a mortality rate, and $B_i(t)$ is the fecundity of species $i$ at time $t$. 
The model postulates that individuals die at rate $\delta$ at each time step, and are replaced by juveniles 
in proportion to the reproductive outputs of the two species that year. Note that $N$ is constant through time. 
Notation for the whole paper is summarized in Table X.
%***DAN: make the table
For each $i$, we assume for simplicity that the random variables $B_i(t)$ are independent and identically 
distributed (iid) through time. In three distinct similations (Fig. \ref{fig:pedagog1}d-f), 
$B_1$ and $B_2$ were, respectively, 
left- (Fig. \ref{fig:pedagog1}a), right- (Fig. \ref{fig:pedagog1}c) and symetrically (Fig. \ref{fig:pedagog1}b)
tail associated, while the same marginal distributions for these fecundities were used in all simulations (see Methods
for details). Thus, the model setup was identical in the three simulations except for different ATAs. 
ATAs substantially influenced aspects of species coexistence in these simulations (Fig. \ref{fig:pedagog1}d-f).

@@

%Goal para
We previously demonstrated that ATAs are common in nature can influence ecological phenomena
\citep{Ghosh_2020_AER, Ghosh_2020_Eco, Ghosh_2020_ecolevol, Ghosh_2021}; and our arguments above
indicate that ATAs may influence species coexistence. Thus, we here seek to quantify the consequences 
of ATAs for competition, coexistence, and the maintenance of biodiversity, by means of the 
following specific goals. 
G1) To augment MCT to show formally how ATAs play a role in coexistence. 
MCT decomposes a measure of coexistence into components due to each of several mechanisms. 
We address how ATAs contribute to these components. 
We examine mechanisms of coexistence that apply in models exhibiting solely temporal variation; 
spatial variation can be considered in future work.
G2) To explore, using the simple lottery model, the relative magnitudes of the contributions to coexistence that come
from ATAs compared to other, previously known mechanisms of coexistence; and to determine whether species may 
sometimes coexist with ATAs but not without, or vice versa. This is to help 
illuminate whether ATA contributions to coexistence and biodiversity maintenance
are likely to be negligible or meaningfully large. 
G3) To provide an empirical example of species coexistence which highlights the role ATAs can play and the
potential importance of ATAs relative to other mechanisms of coexistence.
The example is a diatom system in a chemostat. In the Discussion, we consider whether climate change may alter
ATAs of environmental variables and thereby influence competition and coexistence.
Overall, our study presents a new mechanism of species coexistence
and a means of understanding its theoretical and empirical importance.

@@

We here pursue goal G1 from the Introduction: to augment MCT to show formally how ATAs play a role in coexistence.

\subsection{Theoretical setup}

We develop our new theory for a general two-species model, later applying it to specific scenarios.
Our mathematical notation parallels
that of \cite{Ellner_2016,Ellner_2019}. It is assumed that the growth rate $r_i(t)=\ln(N_i(t+1)/N_i(t))$ 
of species $i$ can be written as an increasing function of
an environment-dependent factor $E_i(t)$ and as a decreasing function of a quantity representing
competitive pressure on species $i$, $C_i(t)$, i.e., $r_i(E_i,C_i)$. We also assume, for simplicity, 
that the $E_i(t)$ are iid through time, and likewise for the
$C_i(t)$. Though the iid assumption can be circumvented, we make the assumption because it simplifies the
presentation of new ideas.

@@

MCT and its recent computational extensions \citep{Ellner_2016,Ellner_2019} 
quantify the contributions of 
multiple mechanisms to coexistence. But we show below
that, for purely temporal variation, only one mechanism relates to tail associations: storage effects. 
So we begin by defining storage effects. 
The storage effect for species $i$ is the contribution of covariation between $E$ and 
$C$ to a difference between rare and common species mean growth rates, when $i$ is rare.
Specifically, storage effects for species $i$ are 
\begin{equation}
\Delta_i^{(EC)} = (\olr_{i\bs i} - q_{ij} \olr_{j \bs i})-(\olrs_{i \bs i}-q_{ij} \olrs_{j \bs i}),\label{eq:storage_effects_gen}
\end{equation}
where: $\olr_{i \bs i}$ refers to the mean ``growth when rare'' (GWR) of species $i$, that is, the mean growth rate of 
species $i$ when it is at negligible abundance and species $j$ (the other species) is at stochastic steady state;
$\olr_{j \bs i}$ refers to the mean steady-state 
growth rate of species $j$ when species $i$ is absent or at negligible abundance (this is $0$, \emph{a priori});
and $\olrs_{i \bs i}$ and $\olrs_{j \bs i}$ are analogous quantities, but for an alternative modelling scenario, identical except that
influences of environment and competition on either species have been rendered independent.  
A subscript 
$i \bs i$ always indicates a quantity computed for species $i$ when that species is at negligible abundance, and 
a subscript $j \bs i$ indicates a quantity computed for $j$ when $i$ is absent.
The \emph{scaling factor}
$q_{ij}$ measures the relative sensitivities to competition of the two species, see \cite{Chesson_1994},
\cite{Ellner_2016} and SI section \ref{SI-SIsect:scaling_factors} for the two
alternative definitions of $q_{ij}$ that we use. 
Since it was assumed that $r_i=r_i(E_i,C_i)$, we can write $\olr_{i \bs i}=\E[r_i(E_i,C_{i \bs i})]$, where
$\E(\cdot)$ denotes expected value just as does an overbar, but we use $\E$ for longer expressions.
Also, $\olr_{j \bs i}=\E[r_j(E_j,C_{j \bs i})]$.
An expression for $\olrs_{i \bs i}$ is obtained from $\olr_{i \bs i}$ by replacing the bivariate random 
variable $(E_i,C_{i \bs i})$ by a random variable with the same marginal distributions, but independent components, 
$(E_i^\shp,C_{i \bs i}^\shp)$, so $\olrs_{i \bs i}=\E[r_i(E_i^\shp,C_{i \bs i}^\shp)]$.
Likewise, $\olrs_{j \bs i}=\E[r_j(E_j^\shp,C_{j \bs i}^\shp)]$.

@@

The other mechanisms in MCT that apply in our modelling context of purely temporal variation
make no reference to covariation between quantities, whereas ATAs are 
defined in a covariation context. Therefore, storage effects are the only mechanism of coexistence
for which ATAs are relevant. Mathematical details supporting this claim are in SI section 
\ref{SI-SIsect:other_mechanisms}, but we here provide a conceptual understanding. \cite{Ellner_2016} and
\cite{Ellner_2019} provide a decomposition of the coexistence metric $\olr_{i \bs i} -q_{ij}\olr_{j \bs i}$,
which equals GWR of species $i$. The decomposition, which we use below, is similar to and motivated by the 
original decomposition of \cite{Chesson_1994}; the decomposition is how the mechanisms of 
coexistence operating in a given case are illuminated. The decomposition is 
\begin{equation}
\olr_{i \bs i} -q_{ij}\olr_{j \bs i}=\Delta_i^0+\Delta_i^E+\Delta_i^C+\Delta_i^{(E\shp C)}+\Delta_i^{(EC)},\label{eq:previous_decomp}
\end{equation}
where: $\Delta_i^0$ is a baseline;
$\Delta_i^E$ is a contribution to coexistence of environmental variation;
$\Delta_i^C$ is a contribution to coexistence of variation in competition;
$\Delta_i^{(E\shp C)}$ is a contribution to coexistence of having simultaneous variation in both environment and competition, but not including the effects 
of covariation in these quantities; and
$\Delta_i^{(EC)}$ is a contribution to coexistence of $EC$ covariation itself -- storage effects. 
Only storage effects involve $EC$ covariation,
so the only mechanism of coexistence influenced by ATA between $E$ and $C$ is storage effects. Combining our decomposition 
(\ref{eq:storage_effects_further}) with the decomposition (\ref{eq:previous_decomp}) of \cite{Ellner_2019} gives 
\begin{equation}
\olr_{i \bs i} -q_{ij}\olr_{j \bs i}=\Delta_i^0+\Delta_i^E+\Delta_i^C+\Delta_i^{(E\shp C)}+\Delta_i^{[E \nat C]} + \Delta_i^{[EC]},\label{eq:full_decomp}
\end{equation}
which is the extension of MCT which was goal G1 of the Introduction.

\subsection{ATA rescue and exclusion}

To evaluate the importance of ATAs for a given scenario,
one can quantify
the magnitude of the new ATA effect term, $\Delta_i^{[EC]}$, relative to the magnitudes of the other terms for the 
scenario.  One typically (but see below) must examine this quantity for each species, $i$,
to obtain complete information about coexistence, since coexistence requires both species populations to 
be able to grow when rare. We also defined scenarios of ``ATA rescue''
and ``ATA exclusion''. \emph{ATA rescue} was considered to have occurred for a given scenario if coexistence
was possible when ATAs were taken into account, i.e., GWR$>0$ for both species, but impossible otherwise, i.e.,
the sum of the non-ATA terms in (\ref{eq:full_decomp}) was negative for at least one species. \emph{ATA exclusion} was 
considered to have occurred if the species did not coexist when ATAs
were taken into account, i.e., GWR$<0$ for at least one species, but would have coexisted were it not for ATAs, 
i.e., the sum of the non-ATA terms in (\ref{eq:full_decomp})
was positive for both species.

@@

\subsection{Versions of the lottery model}

To pursue goal G2 of the Introduction, we applied the framework of Theory to the lottery model, 
comparing the magnitude of the ATA effect on coexistance, $\Delta_i^{[EC]}$, to the magnitudes 
of the other terms of 
the decomposition (\ref{eq:full_decomp}) for various model parameterizations. We used two versions 
of the lottery model which we now specify. The versions both used the same dynamical equations
(Introduction) and assumed temporally independent and identically distributed (iid) fecundities, but the
versions differed in the distribution of $(B_1,B_2)$ used. Both distributions were based on 
transformations of the distributions $(b_1,b_2)$ pictured in Fig. \ref{fig:pedagog1}a-c;
the details of how the left- (Fig. \ref{fig:pedagog1}a), right- (Fig. \ref{fig:pedagog1}c), 
and symmetrically (Fig. \ref{fig:pedagog1}b) tail associated alternatives for $(b_1,b_2)$ were generated 
are in SI section \ref{SI-sect:noise}.

The \emph{log-normal fecundities lottery model} used $B_i = \exp(\sigma+\mu_i b_i)$,
and therefore had parameters $\sigma$, $\mu_1$ and $\mu_2$ in addition to the death rate parameter $\delta$
that both versions of the lottery model use (Introduction). The parameter $\sigma$ is the variance of the
log-fecundities and $\mu_i$ is the mean log fecundity for species $i$.

The \emph{beta-fecundities lottery model}
used $B_i = \eta_i F_\beta \circ \varphi^{-1}(b_i)$, where $\varphi$ is the CDF of the standard normal
distribution and $F_\beta$ is the CDF of a beta distribution 
with shape and scale parameters $0.5$ (probability density function proportional to $x^{-05}(1-x)^{-0.5}$ on
the unit interval). So the beta-fecundities model had parameters $\eta_1$, $\eta_2$ and $\delta$,
and produced fecundities with $U$-shaped distributions between $0$ and $\eta_1$, for $B_1$, and
$\eta_2$, for $B_2$. The mean fecundity for species $i$ was $\eta_i/2$ and larger $\eta_i$ also 
corresponded to more variable fecundities for species $i$.

For both versions of the lottery model, 
fecundities were taken to directly reflect 
the environment, i.e., the $E_i$ of Theory was set equal to $B_i$. 
Competition, $C_i(t)$, in the lottery model occurs at the stage of juveniles occupying open sites.
For both model versions, $C_i$ was therefore taken to be the number of new offspring
divided by the number of open sites, $C_i(t)=( B_1(t)N_1(t)+B_2(t)N_2(t) )/(\delta N$), which
does not depend on $i$. These choices for $E$ and $C$ meet the assumptions of Theory (SI section Y).
%***DAN: This SI reference is to the big section S$ listed below, but this material will be in the
%first part of that section, prior to the subsections.

\subsection{Plotting results for G2}

%Then write a para or section that starts by saying we plotted the decomp terms against model parameters for
%the log-normal-fecundity lottery model, and some mathematical shortcuts simplified it. Then summarize
%the shortcuts.
%This is similar to an existing para. Also cites a lot of sup mat stuff.

%Then a similar para for the beta-fecundity lottery model.

%Then a para about ATA exclusion and rescue and how that was assess for each model - different from
%the current way of doing it with trait spaces - drop the trait space stuff.

%For the sup mat, I think you will need these sections:
%S3 - on generating the Fig. 1 noise (only)
%S4 - as large section with subsections. About the theory for the two lottery models, but not about the effecient computation
%   Starts with material on the general lottery model
%   Then a sub-section with all the computations for the log-normal-fecundities lottery model
%   Then a sub-section for the beta-fecundities lottery model
%S5 - then a similar section, but for the efficient computations
%There might be additional sections, see the current sections and see if those need to be maintained
%as their own sections or can be absorbed into the above.
%This is going to take some work.

\subsection{Diatom chemostat system}

%Then on to the diatom stuff.

### 2023-01-27 paper/SupMat.Rnw
working on revisions to Methods and associated sections of sup mat

@@

We created two cases for $(b_1,b_2)$, left- and right-tail association. 
To generate $M$ points $\left(b_1^{(i)}, b_2^{(i)}\right)$, $i=1,..,M$, for the left tail association case, we first generated M points $(a_1^{(i)},a_2^{(i)})$, $i=1,...,M$, from a bivariate normal distribution $N\left( \vec{0}, \Sigma \right)$, where here $\Sigma = \begin{pmatrix} 1&0\\0&1\end{pmatrix}$. 
Then, for each index $i$, we randomly, with 50\% probability, either (A) let $\tilde b_1^{(i)} = -|a_1^{(i)}|$ and $\tilde b_2^{(i)} = -|a_1^{(i)}|$, or (B) let $\tilde b_1^{(i)} = |a_1^{(i)}|$ and $\tilde b_2^{(i)} = |a_2^{(i)}|$. 
We then let $b_1^{(i)} = \sigma \tilde b_1^{(i)} + \mu_1$ and $b_2^{(i)} = \sigma \tilde b_2^{(i)} + \mu_2$. 
Perfect association in the left tails of the resulting distribution $\left(b_1,b_2\right)$ results from the fact that, in case (A) above, both $\tilde b_1^{(i)}$ and $\tilde b_2^{(i)}$ were assigned to the same value. 
For speed of computation, when samples $(b_1^{(i)},b_2^{(i)})$, $i=1,...,M$ were needed for distinct sets of values of the parameters $\mu_1$, $\mu_2$, and $\sigma$, we used the same $\left(\tilde b_1^{(i)},\tilde b_2^{(i)}\right)$, $i=1,...,M$, simply transforming these values differently (using the different valus of $\mu_1$, $\mu_2$, $\sigma$) for each set of parameter values. Right-tail associated noise was generated in an analogous way.

When needed, noise with symmetric tail association was generated by taking a very large sample (\Sexpr{M}) 
of left-tail associated noise generated as 
described above, and then calculating the Pearson correaltion coefficient, $\rho$, of the sample.
Symmetric noise $\left(\tilde b_1^{(i)},\tilde b_2^{(i)}\right)$ for $i=1,...,M$ was then generated 
from the bivariate normal distribution $N\left(\vec{0}, \Sigma \right)$, where $\Sigma = \begin{pmatrix} 1&\rho \\ \rho&1\end{pmatrix}$, 
and then we set $b_1^{(i)} = \sigma \tilde b_1^{(i)} + \mu_1$ and $b_2^{(i)} = \sigma \tilde b_2^{(i)}+ \mu_2$.

Overall, our version of the lottery model was completely specified by specifying $\delta$, $\mu_1$, $\mu_2$, $\sigma$ and whether 
noise was generated from the symmetric, left- or right-tail associated cases.

>>>>

To generate $M$ points $\left(b_1^{(i)}, b_2^{(i)}\right)$, $i=1,..,M$, for the left-tail association 
case (Fig. \ref{MT-fig:pedagog1}a), we first generated M points $(a_1^{(i)},a_2^{(i)})$, $i=1,...,M$, from a bivariate normal 
distribution $N\left( \vec{0}, \Sigma \right)$, where $\Sigma = \begin{pmatrix} 1&0\\0&1\end{pmatrix}$. 
Then, for each index $i$, we randomly, with 50\% probability, either 
(A) let $b_1^{(i)} = -|a_1^{(i)}|$ and $b_2^{(i)} = -|a_1^{(i)}|$, or 
(B) let $b_1^{(i)} = |a_1^{(i)}|$ and $b_2^{(i)} = |a_2^{(i)}|$. 
Perfect association in the left tails of the resulting distribution $\left(b_1,b_2\right)$ results from 
the fact that, in case (A) above, both $b_1^{(i)}$ and $b_2^{(i)}$ were assigned to the same value. 
Right-tail associated noise (Fig. \ref{MT-fig:pedagog1}c) was generated in an analogous way.

%***DAN: I think I may need the below later, in the computation section, but not here
%For speed of computation, when samples $(b_1^{(i)},b_2^{(i)})$, $i=1,...,M$ were needed for distinct 
%sets of values of the parameters $\mu_1$, $\mu_2$, and $\sigma$, we used the same 
%$\left(\tilde b_1^{(i)},\tilde b_2^{(i)}\right)$, $i=1,...,M$, simply transforming these values 
%differently (using the different valus of $\mu_1$, $\mu_2$, $\sigma$) for each set of parameter values.

When needed, noise with symmetric tail association (Fig. \ref{MT-fig:pedagog1}b) was generated by taking a 
very large sample (\Sexpr{M}) 
of left-tail associated noise generated as 
described above, and then calculating the Pearson correaltion coefficient, 
$\rho$, of the sample.
Symmetric noise $\left(b_1^{(i)},b_2^{(i)}\right)$ for $i=1,...,M$ was then 
generated 
from the bivariate normal distribution $N\left(\vec{0}, \Sigma \right)$, where now
$\Sigma = \begin{pmatrix} 1&\rho \\ \rho&1\end{pmatrix}$.

%***DAN: stopped here on 2023 01 27. See outline in the main text of how I need to restructure the 
%SI sections from this point forward.

@@
